Order reduction and partial MILP models \\ for certifying Robustness of DNNs globally.

### 2025-07-28 AAAI2026/introduction.tex
intro

	\item Since the problem is a very hard verification problem, as the values each neuron can take is no more bound to a small neighborhood, we choose to follow the accurate MILP technique in \cite{ATVA25}. Technically, the number of variables a priori doubles compared with local robustness, from each neuron value in the perturbed image to each neuron value in the perturbed image {\em and} in the original image, as the original image is no more fixed. 
	Recall that the worst case complexity of MILP is exponential in the number of binary variables. A straightforward MILP model would be to use the {\em classical MILP model} \cite{MILP} for each of these variables. While the full MILP model is exact, it is too computationally intensive to be run for anything but small DNNs. As explained in \cite{ATVA25}, many variables need to be linearly relaxed 
	to reach useable bounds. The main issue with the classical model is that its linear relaxations is extremely inaccurate, as the variables are independent of each other. Instead, we develop another exact MILP model,
	called the {\em "2v" model}, where the variables are the values of the perturbed neurons, as well as the difference between the original and the perturbed neuron values (called the {\em diff variables}). We study and encode exactly how the diff variables evolve after passing through a ReLU (figure \ref{fig.2v}). In this model, the linear relaxation is much more accurate, as each 
	{\em diff variable} can be bounded 	after a ReLU as a function of the value of the {\em diff variables} before the ReLU. We propose 2 lighter models, namely the {\em "3v" model} which decouples one of the variable (less constraint means less accuracy), as well as an efficient {\em "1v" model} only considering the {\em diff variables} but forgetting the value of the perturbed neurons (with half the binary variables).

	\item We adapt the Solution Aware Scoring from \cite{ATVA25} to our novel MILP models, in order to select the most important ReLUs to be treated using complex binary variables, while less important variables are treated using linear relaxation.

	\item Results obtained on the full input space have several shortcomings: while the bounds are fully correct, they are also particularly pessimistic, as all inputs, including absolutely improbable ones very far away from anything the network has been trained on, need to be accounted for. Also, the runtime to obtain the bound is particularly long. Finally, when asking for worst-case 
	scenarios, improbable images are generated. To solve this problem, we restrict the input space to ignore all these improbable inputs, using model order reduction techniques from engineerial science \cite{aiware}.

\end{enumerate}

>>>>

	\item Since the problem is a very hard verification problem, as the values each neuron can take is no more bound to a small neighborhood, we choose to follow the accurate MILP technique in \cite{ATVA25}. Technically, the number of variables a priori doubles compared with local robustness, from each neuron value in the perturbed image to each neuron value in the perturbed image {\em and} in the original image, as the original image is no more fixed. 
	Recall that the worst case complexity of MILP is exponential in the number of binary variables \cite{DivideAndSlide}. A straightforward MILP model would be to use the {\em classical MILP model} \cite{MILP} for each of these variables. While the full MILP model is exact, it is too computationally intensive to be run for anything but small DNNs. As explained in \cite{ATVA25}, many variables need to be linearly relaxed to reach useable bounds. The main issue with the classical model is that its linear relaxations is extremely inaccurate, as the variables are independent of each other. Instead, we develop another exact MILP model, called the {\em "2v" model}, where the variables are the values of the perturbed neurons, as well as the difference between the original and the perturbed neuron values (called the {\em diff variables}). We study and encode exactly how the diff variables evolve after passing through a ReLU (figure \ref{fig.2v}). In this model, the linear relaxation is much more accurate, as each 
	{\em diff variable} can be bounded 	after a ReLU as a function of the value of the {\em diff variables} before the ReLU. We propose 2 lighter models, namely the {\em "3v" model} which decouples one of the variable (less constraint means less accuracy), as well as an efficient {\em "1v" model} only considering the {\em diff variables} but forgetting the value of the perturbed neurons (with half the binary variables).

	\item We adapt the Solution Aware Scoring from \cite{ATVA25} to our novel MILP models, in order to select the most important ReLUs to be treated using complex binary variables, while less important variables are treated using linear relaxation. The chosen number of binary variables depends upon the complexity of the DNN as well as the targeted runtime.

   \item  In terms of perturbations, we consider conjunctions of $L_0$ and $L_1$ norms, which allows to accurately describe perturbations. For instance, "each subpixel is perturbed by at most $\frac{20}{255}$ ($L_0$) and the sum of the absolute value of perturbations over all subpixels is at most $\frac{150}{255}$" ($L_1$ perturbation). While $L_1$ perturbations are not linear (because of the absolute values), reason for which it is seldom used, we explain how to use it as a perturbation in the MILP model without incurring any expansive binary variables (only cheap linear variables are necessary to encode $L1$ perturbations).

	\item Results obtained on the full input space have several shortcomings: while the bounds are fully correct, they are also particularly pessimistic, as all inputs, including absolutely improbable ones very far away from anything the network has been trained on, need to be accounted for. Also, the runtime to obtain the bound is particularly long. Finally, when computing worst-case 
	pairs (image, perturbation), improbable images are generated, hence these  are not meaningful. To solve this problem, we restrict the input space to ignore improbable inputs, using model order reduction techniques from engineering science \cite{aiware}. Specifically, we use Principal Component Analysis (PCA) to reduce the space to a linear input space. We choose the number of dimensions of the space to equal the accuracy of the DNN on the reduced space. 
	%(using a projection to the reduced space then the inverse projection to obtain a very similar image understandable by the DNN). 
	On the MNIST benchmark, this means 20 linear dimensions to match the $97\%$ accuracy of the DNN we considered, instead of the 768 dimensions of the full image space. Using PCA, which is linear, makes it easy to specify perturbations on the actual image (where it is meaningful) rather than on the reduced space.

\item Experimentally, this allows to obtain bounds for a considered pre-trained DNN for MNIST (with accuracy $97\%$) that can certify in real time that $86\%$ of the images are robust for an $L_1$ perturbation of $0.5$, resp. $53\%$ for an $L_1$ perturbation of $1$. We finally experimented our MILP models for {\em regression} tasks, considering a DNN surrogate built from a slow physical finite element model, providing the deformation and the plastic strain of a pipe \cite{aiware}. The DNN has been learnt on a reduced order basis already of dimension 10, from thousands of input dimensions. For regression, the aim is to understand if the DNN surrogate that has been learnt is brittle to small perturbations, which could make it unfaithful to the physical model. Here we want to obtain bounds (which is what our models provide), as well as generating worst-case (image,perturbation) pair close to the bound, where the strain is particularly different, that can be run in the physical model to understand whether the DNN surrogate is faithful. Our tool was helpful to provide limited bounds, as well as almost matching worst-case (image,perturbation) pair that was confirmed in the physical model.

\item In both cases, the best bounds were obtained using the "3v" model, as the theoretically more accurate "2v" model did not converge enough even after a very large timeout. Still, when trying to produce worst case pairs (image, perturbation), which is somehow easier than certifying bounds as heuristics to guide towards a solution can be used, the "2v" model found worser-case solutions. Last, the "1v" model produced bounds as good as the "2v" models and much faster than the "3v" models, although slightly larger. That is, all 3 models we propose have favorable trade-offs in different contexts.
\end{enumerate}
